% ============================================================
% Abstract
% BearMamba-3: Mamba-3 + L_kin + Multi-Sensor Fusion
% Target: IEEE TIM (IF 5.6)  Draft: 2026-06-15 → 2026-06-18 (D34)
% ============================================================

\begin{abstract}
Reliable diagnostic instrumentation for rolling bearings demands accurate
and interpretable inference from vibration measurements.
State-space models (SSMs) have recently achieved competitive accuracy on
vibration-based bearing fault diagnosis, yet embedding physical domain
knowledge at the activation level remains largely unexplored.
This paper presents \textbf{BearMamba-3}, a bearing diagnostic framework
built on the official Mamba-3 backbone (ICLR~2026), augmented with a
\textit{kinematic frequency inductive bias} loss $\mathcal{L}_{\text{kin}}$
and a multi-sensor convolutional input pathway.

$\mathcal{L}_{\text{kin}}$ exploits a property unique to Mamba-3's complex
oscillatory states: each state dimension carries an instantaneous rotation
frequency $f_{h,k}(t)$, interpretable as a physical resonance in Hz.
The loss aligns these state frequencies with fault harmonics computed per
sample from the shaft speed — a constraint undefined for real-valued
Mamba-2, structurally binding the Mamba-3 backbone to the kinematic
loss design.

Experiments across CWRU, Paderborn University (PU), and XJTU-SY yield
four findings:
\textit{(1)} BearMamba-3 achieves accuracy non-inferior to BearMamba-2
($p>0.05$, $n=5$, cross-condition PU), confirming Mamba-3 does not trade
accuracy for interpretability;
\textit{(2)} $\mathcal{L}_{\text{kin}}$ reduces inter-seed variance by
14--62\% under AWGN on CWRU (descriptive; $p\geq0.25$), and frequency
snapshots visualise progressive fault-harmonic alignment during training;
\textit{(3)} dual-sensor fusion yields a statistically significant
$+0.70$~pp gain at SNR~$=-4$~dB on CWRU (Wilcoxon $p=0.031$, $n=5$),
while cross-condition inner-race recall improves $+31.6$~pp on XJTU-SY
(5/5 seeds, $p=0.063$), characterising a precision--sample-efficiency
trade-off;
\textit{(4)} a cross-dataset BatchNorm ablation reveals that 1D-CNN's
within-condition advantage over BM3 is principally attributable to
BatchNorm regularisation (removing BatchNorm reduces 1D-CNN by $12.7$~pp
at $-8$~dB, $p=0.063$; without BatchNorm, $p=0.125$ vs BM3), whereas
BM3~$+\mathcal{L}_{\text{kin}}$ achieves $+38.4$~pp macro-F1 on
XJTU-SY cross-condition evaluation (5/5 seeds, $p=0.063$), demonstrating
that physical frequency alignment provides out-of-distribution robustness
that statistical normalisation cannot replicate.
Code and configurations are publicly available.

\end{abstract}

\begin{IEEEkeywords}
Bearing fault diagnosis,
state-space model,
Mamba-3,
physical inductive bias,
kinematic loss,
multi-sensor fusion,
instantaneous frequency,
vibration measurement,
diagnostic instrumentation.
\end{IEEEkeywords}
