% ============================================================
% Section 1: Introduction
% BearMamba-3: Mamba-3 + L_kin + Multi-Sensor Fusion
% Target: IEEE TIM (IF 5.6)  Draft: 2026-06-15 → 2026-06-18 (D34)
% ============================================================

\section{Introduction}
\label{sec:intro}

Rolling element bearings are among the most failure-prone components in
rotating machinery; their unexpected breakdown accounts for a disproportionate
share of unplanned industrial downtime~\cite{bearing_importance}.
Reliable on-line vibration measurement and condition monitoring are
therefore central to predictive maintenance instrumentation, and
data-driven fault diagnosis---the automatic identification of fault
location (inner race, outer race, rolling element, cage) from vibration
signals---has attracted sustained research attention over the past
decade~\cite{lei2020applications}.

\paragraph{From CNNs to sequence models.}
Early deep-learning methods relied on 1-D and 2-D convolutional neural
networks that treat a sliding vibration window as a fixed-length pattern
classification problem~\cite{wen2018new,zhang2020deep}.
Transformers subsequently demonstrated that long-range temporal modelling
can capture fault-periodic harmonics more faithfully than local
convolutions~\cite{shao2021transformer_bearing},
but their quadratic self-attention cost scales poorly with window length.
State-space models (SSMs) offer a compelling middle ground:
linear-time sequence processing with a principled recurrent mechanism
that can, in principle, encode periodicity through its transition
eigenstructure~\cite{gu2021efficiently}.
Our prior work introduced BearMamba-2, an SSM bearing diagnostic model
using the Mamba-2 backbone that achieves competitive accuracy under AWGN
and cross-condition transfer, and showed that a complex-state extension
(KG-cSSM) further benefits variable-speed diagnosis through
architecture-level physics priors.

\paragraph{Gap: activation-level physical alignment for Mamba-3.}
Recent work has embedded physical priors into bearing diagnostics at
different levels of the model: Jiang et al.~\cite{jiang2025mechanism} guide
feature alignment via mechanism-driven domain adaptation at the
distribution level; PINNs~\cite{raissi2019physics,willard2022physics}
impose governing equations on network outputs.
However, none of these approaches can leverage the Mamba-3 architecture's
unique capability: \emph{per-step oscillatory states} that carry instantaneous
physical frequencies as first-class activation values.
The Mamba-3 architecture~\cite{mamba3_2026} parameterises each state
dimension by an instantaneous angle activation $\theta_{h,k}(t)$
projected from the current input, yielding a sequence of instantaneous
physical frequencies $f_{h,k}(t)$ (measured in Hz) interpretable as the
momentary resonance frequency of each state at each time step.
We observe that $f_{h,k}(t)$ provides a direct \emph{activation-level hook}
for injecting bearing kinematics: by encouraging the model's internal
oscillation frequencies to align with the analytically known fault harmonics
($f_{\text{BPFO}}$, $f_{\text{BPFI}}$, BSF, and their harmonics —
derived in Section~\ref{sec:physics}), we endow the model with a
physically meaningful inductive bias without modifying its architecture
or adding parameters at inference time.

\paragraph{Gap: multi-sensor Mamba-3.}
Practical deployment often provides more than one vibration channel:
drive-end and fan-end accelerometers in lab rigs, or orthogonal
horizontal and vertical axes in industrial field sensors.
Multi-sensor fusion at the convolutional embedding stage is architecturally
trivial but its \emph{interaction} with the SSM state bandwidth
($\texttt{mimo\_rank}$ in Mamba-3) has not been characterised.
Understanding when additional sensor channels improve diagnosis — and when
they induce negative transfer in low-data regimes — is practically important.

\paragraph{Contributions.}
This paper proposes \textbf{BearMamba-3}, a bearing fault diagnosis framework
with the following four contributions:

\begin{enumerate}[leftmargin=*, label=\textbf{C\arabic*}]
  \item \textbf{First Mamba-3 bearing diagnostic model with structural
        binding to physical frequency alignment.}
        We demonstrate that the ICLR~2026 official Mamba-3
        implementation~\cite{mamba3_2026} can be effectively applied to
        bearing diagnosis, and—crucially—that its oscillatory state mechanism
        is the \emph{unique carrier} of our kinematic loss $\mathcal{L}_{\text{kin}}$:
        real-valued SSMs (Mamba-2) lack instantaneous rotation frequencies,
        rendering $\mathcal{L}_{\text{kin}}$ architecturally undefined for them.
        The adoption of Mamba-3 and the design of $\mathcal{L}_{\text{kin}}$ are
        therefore a single integrated decision (Section~\ref{subsec:lkin}).

  \item \textbf{$\mathcal{L}_{\text{kin}}$: activation-level kinematic inductive bias.}
        We formulate a lightweight auxiliary loss that penalises the distance
        between each model state's instantaneous frequency and the nearest
        bearing fault harmonic, computed per sample from the actual shaft
        speed.
        $\mathcal{L}_{\text{kin}}$ requires no additional parameters, is
        computed only during training, and naturally accommodates variable-speed
        signals via per-sample RPM targets.
        Empirically, $\mathcal{L}_{\text{kin}}$:
        \textit{(i)} improves physical interpretability by aligning state
        frequencies with fault harmonics, quantified through I1 snapshots
        (Section~\ref{subsubsec:i1});
        \textit{(ii)} reduces inter-seed accuracy variance under AWGN
        (std $\downarrow$14--62\% across the EAAI evaluation SNR grid
        for CWRU; Section~\ref{subsubsec:snr_ablation}), a descriptive
        observation without statistical significance claim at $\alpha=0.05$;
        \textit{(iii)} reveals the boundary conditions of activation-level
        physics priors: ineffective in discrete-speed regimes (Paderborn) and
        amplitude-dominated run-to-failure signals (XJTU-SY).
        % ⚠️ std 范围 14-62%（不得写"30-63%"），D15

  \item \textbf{Multi-sensor fusion with precision--sample-efficiency
        trade-off characterisation.}
        We extend BearMamba-3 to multi-sensor input via a shared convolutional
        embedding ($n_s$ input channels) and ablate the MIMO--sensor-count
        interaction through systematic experiments on CWRU (drive-end + fan-end)
        and XJTU-SY (horizontal + vertical).
        \textbf{The sole statistically significant finding} is a dual-sensor
        accuracy gain of $+0.70$~pp at SNR~$= -4$~dB on CWRU
        (Wilcoxon $p = 0.031$, $n = 5$, one-sided).
        XJTU-SY experiments provide complementary mechanistic evidence of a
        precision--sample-efficiency trade-off: dual-sensor cross-condition
        inner-race recall improves by $+31.6$~pp (5/5 seeds concordant,
        $p = 0.063$, above $\alpha=0.05$), while leave-one-bearing-out with
        a single OR training instance induces negative transfer ($-15.5$~pp,
        $p = 0.063$, 5/5 seeds negative).
        The XJTU results are reported as mechanistic evidence, not as
        statistically significant claims (Section~\ref{subsubsec:xjtu}).
        % ⚠️ 不声称"XJTU统计显著"（D29）；统计主证据=CWRU -4dB p=0.031

  \item \textbf{Cross-paradigm analysis: BN-dependent vs.\ frequency-grounded
        inductive bias.}
        A systematic BatchNorm ablation on 1D-CNN reveals that BatchNorm is the
        principal source of 1D-CNN's within-condition low-SNR advantage
        ($-12.7$~pp without BatchNorm at $-8$~dB, $p=0.063$; gap with BM3
        becomes non-significant, $p=0.125$).
        Concurrently, BearMamba-3~$+\mathcal{L}_{\text{kin}}$ achieves
        $+38.4$~pp macro-F1 on XJTU-SY cross-condition OOD evaluation
        (5/5 seeds, $p=0.063$) over 1D-CNN, which collapses to a
        constant-OR predictor.
        This exposes a systematic complementarity: the \emph{BN-dependent
        paradigm} confers powerful within-distribution regularisation at the
        cost of OOD brittleness; the \emph{frequency-grounded paradigm}
        provides distribution-agnostic physical constraints at no within-condition
        cost once BatchNorm is absent.
        (Section~\ref{subsec:paradigm})
        % ⚠️ D33: Win2 叙事框架最终版；禁用"BM3 significantly outperforms CNN-noBN"
\end{enumerate}

\paragraph{Relationship to prior work.}
BearMamba-3 is positioned in a progression of increasingly deep physics
integration: feature-level noise-awareness (BearMamba-2, submitted) $\to$
architecture-level complex-state priors for variable-speed diagnosis
(KG-cSSM, under review) $\to$
\textbf{BearMamba-3} (activation-level kinematic inductive bias $\times$
multi-sensor).
BearMamba-3 is differentiated from the KG-cSSM line by using the official
Mamba-3 backbone~\cite{mamba3_2026} (rather than a custom complex-state
design), by per-sample variable-speed loss computation, and by the
systematic multi-sensor ablation.

\paragraph{Experimental scope.}
All experiments use 5 independent random seeds with Wilcoxon signed-rank
tests ($n=5$).
On CWRU, we benchmark against a classical baseline
(SK-SVM~\cite{antoni2006spectral_kurtosis}), two deep learning baselines
(1D-CNN and Transformer-1D), and BearMamba-2 (Table~\ref{tab:cwru_baselines});
focused ablation studies then isolate the Mamba-3 vs Mamba-2 and
$\pm\mathcal{L}_{\text{kin}}$ effects across the EAAI SNR evaluation grid
($-8$ to $+10$~dB).

\paragraph{Paper organisation.}
Section~\ref{sec:related} reviews related work.
Section~\ref{sec:physics} provides the bearing kinematics background and
defines the fault characteristic frequency equations used throughout.
Section~\ref{sec:method} describes the BearMamba-3 architecture and
$\mathcal{L}_{\text{kin}}$.
Section~\ref{sec:experiments} presents experimental results on CWRU,
Paderborn, and XJTU-SY.
Section~\ref{sec:discussion} discusses findings, boundary conditions, and
limitations.
Section~\ref{sec:conclusion} concludes.
